---
title: "Data Preparation and Feature Engineering"
subtitle: "Home Credit Default Risk - Capstone Project"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
    number-sections: true
execute:
  warning: false
  message: false
---

# Overview

This document describes the data cleaning, feature engineering, and aggregation pipeline for the Home Credit Default Risk competition. The goal is to transform raw application data and supplementary datasets into a clean, feature-rich dataset ready for modeling.

## Objectives

1. **Clean and transform** application data based on EDA findings
2. **Create engineered features** including demographic transformations and financial ratios
3. **Aggregate supplementary data** (bureau, previous applications, installments) to applicant level
4. **Ensure train/test consistency** by fitting parameters on training data only

## Data Sources

- `application_train.csv` / `application_test.csv`: Main application data
- `bureau.csv`: Credit Bureau data on previous credits
- `previous_application.csv`: Previous loan applications at Home Credit
- `installments_payments.csv`: Payment history for previous loans

```{r}
#| label: setup
#| message: false

# Load required libraries
library(tidyverse)
library(janitor)
library(knitr)
library(scales)

# Source the preprocessing functions
source("home_credit_preprocessing.R")
```

# Data Loading

First, let's load the raw datasets to understand what we're working with.

```{r}
#| label: load-data

# Load main application data
train_raw <- read_csv("application_train.csv", show_col_types = FALSE)
test_raw <- read_csv("application_test.csv", show_col_types = FALSE)

# Display dimensions
tibble(
  Dataset = c("Training", "Test"),
  Rows = c(nrow(train_raw), nrow(test_raw)),
  Columns = c(ncol(train_raw), ncol(test_raw))
) |> kable()
```

# Data Cleaning

## DAYS_EMPLOYED Anomaly

During EDA, we identified that `DAYS_EMPLOYED` contains the placeholder value 365243 (~1000 years), representing unemployed/pensioners.

```{r}
#| label: days-employed-anomaly

# Check the anomaly
train_raw |>
  count(days_employed_anomaly = DAYS_EMPLOYED == 365243) |>
  mutate(pct = percent(n / sum(n)))
```

The `fix_days_employed_anomaly()` function handles this by:

1. Creating a binary indicator `days_employed_anomaly` (missingness may be predictive)
2. Replacing the placeholder with NA

## EXT_SOURCE Missing Values

The `EXT_SOURCE` variables are highly predictive external credit scores with significant missingness:

```{r}
#| label: ext-source-missing

train_raw |>
  summarise(
    ext_source_1_missing = percent(mean(is.na(EXT_SOURCE_1))),
    ext_source_2_missing = percent(mean(is.na(EXT_SOURCE_2))),
    ext_source_3_missing = percent(mean(is.na(EXT_SOURCE_3)))
  ) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Pct")
```

The `impute_ext_source()` function imputes these using **median values computed from training data only** to prevent data leakage.

## Other Cleaning Steps

The `clean_application_data()` function also handles:

- Converting "XNA" strings to NA (found in `ORGANIZATION_TYPE`)
- Standardizing column names to snake_case
- Ensuring flag columns are integers

# Feature Engineering

## Demographic Features

The `create_demographic_features()` function transforms negative day values and creates age-related features:

| Feature | Description |
|---------|-------------|
| `age_years` | Age converted from negative days to positive years |
| `employment_years` | Employment duration in years |
| `registration_years` | How long client has been registered |
| `employment_to_age_ratio` | Fraction of adult life employed |
| `age_group` | Binned age categories |
| `is_working_age` | Flag for ages 18-65 |
| `is_young_adult` | Flag for ages less than 30 |

## Financial Ratios

The `create_financial_ratios()` function computes common credit risk ratios:

### Core Credit Ratios

| Ratio | Formula | Interpretation |
|-------|---------|----------------|
| Credit-to-Income | AMT_CREDIT / AMT_INCOME_TOTAL | Higher = more leveraged |
| Annuity-to-Income (DTI proxy) | AMT_ANNUITY / AMT_INCOME_TOTAL | Payment burden |
| Credit-to-Goods (LTV proxy) | AMT_CREDIT / AMT_GOODS_PRICE | Greater than 1 means financing more than goods value |

### Additional Financial Metrics

| Feature | Description |
|---------|-------------|
| `credit_term_months` | Loan term derived from credit/annuity |
| `income_per_person` | Income divided by family members |
| `down_payment_ratio` | Proxy for down payment / goods price |
| `ext_source_mean` | Average of external scores |
| `ext_source_weighted` | Weighted average (emphasizing more predictive scores) |

## Missing Data Indicators

The `create_missing_indicators()` function adds binary flags for key variables with missingness, as missing patterns are often predictive:

| Variable | Typical Missingness |
|----------|---------------------|
| ext_source_1 | ~56 percent |
| ext_source_3 | ~20 percent |
| occupation_type | ~31 percent |
| own_car_age | ~66 percent |
| amt_goods_price | ~3 percent |

## Interaction and Binned Features

The `create_interaction_features()` function creates:
 
- **Binned variables**: Income and credit quintiles (thresholds from training data)
- **Categorical interactions**: Gender x Car ownership, Education x Income bin
- **Numeric interactions**: Age x log(Income), External score x log(Credit)
- **Polynomial features**: Squared terms for external scores and age
- **Aggregates**: Document count, contact method count

# Supplementary Data Aggregation

## Bureau Data

The `aggregate_bureau()` function creates features from Credit Bureau history:

| Category | Features |
|----------|----------|
| Credit Counts | Total credits, active/closed/sold/bad debt counts, active ratio |
| Amounts | Total/mean/max credit amounts, debt amounts, utilization ratio |
| Overdue | Overdue counts, amounts, max days past due, DPD ratio |
| Timing | Days since first/last credit, credit duration |

## Previous Applications

The `aggregate_previous_application()` function creates features from prior Home Credit applications:

| Category | Features |
|----------|----------|
| Application Counts | Total, approved/refused/canceled counts |
| Rates | Approval rate, refusal rate |
| Amounts | Applied/credited sums and means, credit-to-application ratio |
| Refusal Reasons | Client-related refusals, scoring-related refusals |

## Installments Payments

The `aggregate_installments()` function creates payment behavior features:

| Category | Features |
|----------|----------|
| Late Payments | Late count/rate, very late (30+ days) count/rate |
| Days Late | Mean/max/sum days late |
| Payment Amounts | Total paid vs expected, payment ratio, underpayment metrics |
| Trends | Version counts (indicates payment plan changes) |

# Train/Test Consistency

## Parameter Fitting Strategy

To prevent **data leakage**, all transformations that require statistical computation use parameters fitted on training data only:

1. **Medians for imputation**: Computed from training, applied to both
2. **Bin thresholds**: Quintile breaks computed from training
3. **Column alignment**: Ensure test has same columns as train (except TARGET)

The `fit_preprocessing_params()` function stores all parameters needed for consistent transformation.

## Column Consistency

The `ensure_column_consistency()` function:

1. Adds missing columns to test (filled with NA)
2. Removes extra columns from test
3. Reorders test columns to match training

# Running the Pipeline

## Full Pipeline Execution

```{r}
#| label: run-pipeline
#| cache: true

# Run the complete preprocessing pipeline
result <- run_preprocessing_pipeline(
  train_path = "application_train.csv",
  test_path = "application_test.csv",
  bureau_path = "bureau.csv",
  prev_app_path = "previous_application.csv",
  installments_path = "installments_payments.csv"
)

# Extract results
train_processed <- result$train
test_processed <- result$test
params <- result$params
```

## Pipeline Results Summary

```{r}
#| label: pipeline-summary

# Summary of processed data
tibble(
  Dataset = c("Training", "Test"),
  Rows = c(nrow(train_processed), nrow(test_processed)),
  Columns = c(ncol(train_processed), ncol(test_processed))
) |> kable(caption = "Processed Dataset Dimensions")
```

### New Features Created

```{r}
#| label: feature-count

# Count features by category
original_cols <- ncol(train_raw)
final_cols <- ncol(train_processed)

tibble(
  Category = c("Original features", "New features", "Total"),
  Count = c(original_cols, final_cols - original_cols, final_cols)
) |> kable()
```

### Feature Categories

```{r}
#| label: feature-categories

# Categorize features
feature_names <- names(train_processed)

tibble(
  Category = c(
    "Bureau aggregations",
    "Previous application aggregations",
    "Installments aggregations",
    "Missing indicators",
    "Financial ratios",
    "Demographic features",
    "Interactions"
  ),
  Count = c(
    sum(str_starts(feature_names, "bureau_")),
    sum(str_starts(feature_names, "prev_app_")),
    sum(str_starts(feature_names, "inst_")),
    sum(str_starts(feature_names, "missing_")),
    sum(str_detect(feature_names, "ratio|to_income|to_credit|per_")),
    sum(str_detect(feature_names, "age_|employment_|years")),
    sum(str_detect(feature_names, "interaction|_bin$|squared"))
  )
) |> kable()
```

# Data Quality Checks

## Verify No Data Leakage

```{r}
#| label: verify-no-leakage

# Check that TARGET only exists in training data
tibble(
  Check = c(
    "TARGET in train",
    "TARGET in test",
    "Same columns (except TARGET)"
  ),
  Result = c(
    "target" %in% names(train_processed),
    "target" %in% names(test_processed),
    setequal(
      setdiff(names(train_processed), "target"),
      names(test_processed)
    )
  )
) |> kable()
```

## Missing Value Summary

```{r}
#| label: missing-summary

# Calculate missing percentages for key features
train_processed |>
  select(
    ext_source_1, ext_source_2, ext_source_3,
    employment_years, credit_to_income,
    bureau_credit_count, prev_app_count, inst_payment_count
  ) |>
  summarise(across(everything(), \(x) percent(mean(is.na(x))))) |>
  pivot_longer(everything(), names_to = "Feature", values_to = "Missing_Pct") |>
  kable()
```

# Save Processed Data

```{r}
#| label: save-data

# Save processed datasets
write_csv(train_processed, "train_processed.csv")
write_csv(test_processed, "test_processed.csv")

# Save preprocessing parameters for reproducibility
saveRDS(params, "preprocessing_params.rds")

message("Processed data saved successfully!")
```

# Function Reference

## Cleaning Functions

| Function | Purpose |
|----------|---------|
| `fix_days_employed_anomaly()` | Replace 365243 placeholder with NA |
| `impute_ext_source()` | Median imputation for EXT_SOURCE variables |
| `clean_application_data()` | Master cleaning function |

## Feature Engineering Functions

| Function | Purpose |
|----------|---------|
| `create_demographic_features()` | Age, employment duration, age groups |
| `create_financial_ratios()` | Credit ratios, income ratios, score combinations |
| `create_missing_indicators()` | Binary flags for missing values |
| `create_interaction_features()` | Interactions, bins, polynomial terms |

## Aggregation Functions

| Function | Purpose |
|----------|---------|
| `aggregate_bureau()` | Summarize Credit Bureau history |
| `aggregate_previous_application()` | Summarize prior Home Credit applications |
| `aggregate_installments()` | Summarize payment behavior |

## Pipeline Functions

| Function | Purpose |
|----------|---------|
| `fit_preprocessing_params()` | Compute parameters from training data |
| `apply_preprocessing()` | Apply transformations using stored params |
| `join_all_features()` | Join aggregated features to main data |
| `ensure_column_consistency()` | Align train/test columns |
| `run_preprocessing_pipeline()` | Execute complete pipeline |

# Session Info

```{r}
#| label: session-info

sessionInfo()
```